On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-22 main.tex
some fixes to the intro

\paragraph{Notation:} In this paper, we will use un-curried notation to apply curried functions, for example, given a function \(δ: X → Y → Z\), we will write the function applications as \(δ(x): Y → Z\) and \(δ(x, y): Z\).

\section{Background on Coalgebra and GKAT}

>>>>

Kleene Algebra with Tests (KAT) is a staple algebra system to reason about program equivalence, and have found applications in numerous fields, including software-defined networks~\cite{anderson_NetKATSemanticFoundations_2014a,moeller_KATchFastSymbolic_2024,smolka_ScalableVerificationProbabilistic_2019,smolka_CantorMeetsScott_2017}, program transformations~\cite{kozen_BohmJacopiniTheorem_2008,angus_KleeneAlgebraTests_2001,kozen_NonlocalFlowControl_2008,grathwohl_KAT_2014,kozen_KleeneAlgebraTests_1996}, and program logics~\cite{antonopoulos_AlgebraAlignmentRelational_2023,zhang_DomainReasoningTopKAT_2024,zhang_IncorrectnessLogicKleene_2022c,kozen_HoareLogicKleene_2000,desharnais_ModalKleeneAlgebra_2004}.
Beyond its theoretical achievements, Kleene Algebra with Tests has also inspired several practical tools for specialized domains.
Interestingly, despite the PSPACE-complete complexity of KAT, many tools developed are able to match, or even exceed, the performance of specialized state-of-the-art tools in their respective domains~\cite{moeller_KATchFastSymbolic_2024,smolka_ScalableVerificationProbabilistic_2019}.

This surprising fact lead to the development of Guarded Kleene Algebra with Tests (GKAT)~\cite{smolka_GuardedKleeneAlgebra_2020}, a (co)algebraic system that seeks to encapsulate this efficient fragment of KAT.
Indeed, the determinism of GKAT gives it a different coalgebra structure from KAT, and allows for a GKAT expression to be directly converted to a linear-sized coalgebra, as opposed to the exponential sized coalgebra produced KAT in general~\cite{smolka_GuardedKleeneAlgebra_2020,kozen_CoalgebraicTheoryKleene_2017}.

However, as it current stands, the desirable theoretical properties of GKAT do not translate to tangible real-world performance. 
The decision procedures for GKAT can often be slower than that of KAT, primarily due to the following two reasons.

First, the natural semantics of GKAT coalgebra is not the finite-trace model, but the infinite-trace model~\cite{smolka_GuardedKleeneAlgebra_2020,schmid_GuardedKleeneAlgebra_2021}.
Thus, to determine finite-trace equivalence, it is necessary to \emph{normalize} the coalgebra first, requiring one to compute the entire coalgebra ahead-of-time; and iterate through its entirety.
Without the need to normalize prior to computing bisimulation, the decision procedure of KAT equivalence can terminate as soon as a counter-example has been found which can lead to better performance than the original GKAT decision algorithm~\cite{smolka_GuardedKleeneAlgebra_2020}.

The lack of symbolic algorithms also makes GKAT a less desirable foundation for performance-critical applications.
Although the generated coalgebra from GKAT expression have a linear number of states, the number of transitions is exponential to \emph{all} the primitive tests appearing in the entire expression.
This inefficiency is noted in previous literature, and poses a real limitation on real-world systems~\cite{smolka_GuardedKleeneAlgebra_2020,zhang_CFGKATEfficientValidation_2025}.
On the other hand, symbolic algorithms for KAT have been proven to be effective for generic equality solving and domain-specific applications~\cite{pous_SymbolicAlgorithmsLanguage_2015,moeller_KATchFastSymbolic_2024}.

In this paper, we present two novel algorithms that address both shortcomings, and demonstrate how the coalgebra theory of GKAT serves as the foundation for the design and correctness of these algorithms.

The first algorithm supports on-the-fly coalgebra generation by performing bisimulation in a greedy manner.
Specifically, the liveness checking of states only kicks in when the bisimulation algorithm encounters a disparity.
We show the correctness of this algorithm through a novel coalgebraic concept which we named \emph{greedy bisimulation}.

The second algorithm lifts the first into a symbolic setting, where our development is based on an extension of GKAT coalgebra, which we named \emph{symbolic GKAT coalgebra}.
In contrast to the symbolic KAT derivative~\cite{pous_SymbolicAlgorithmsLanguage_2015}, the deterministic nature of GKAT coalgebra allows its symbolic variant to transition via boolean expressions instead of primitive tests. 
This optimization not only allows more compact coalgebra representation, but also allows the usage of arbitrary off-the-shelf SAT solvers to resolve boolean constraints, instead of being limited to binary decision diagrams as utilized by the symbolic algorithm of KAT~\cite{pous_SymbolicAlgorithmsLanguage_2015}. 
Finally, we lift both the derivative construction~\cite{schmid_GuardedKleeneAlgebra_2021} and Thompson's construction~\cite{smolka_GuardedKleeneAlgebra_2020} to the symbolic setting, and prove their correspondence, correctness, and finiteness.

To evaluate the practical impact of our work, we implement our algorithms in Rust. We show through experiments that
our implementation performs orders of magnitude better than existing implementation for KAT~\cite{pous_SymbolicAlgorithmsLanguage_2015},
confirming that the theoretical elegance of GKAT indeed leads to significant real-world performance improvements.

We list our contributions and structures of this paper as follows:
\begin{itemize}
    \item In~\Cref{sec:background}, we recap the necessary concepts in universal coalgebra, and the coalgebraic theory of GKAT. In particular, we do not define the semantics of GKAT either algebraically or coalgebraically, as we can avoid directly interacting with the semantics by leveraging the coalgebraic theory of GKAT.
    \item In~\Cref{sec:greedy-bisim}, we define \emph{greedy bisimulation} and developed its coalgebraic theory to serve as the ground for the correctness of our algorithm.
    We also performs a brief complexity analysis, and argue that our algorithm is almost always more efficient than the original normalize-bisimulate approach~\cite{smolka_GuardedKleeneAlgebra_2020}.
    \item In~\Cref{sec:symb-gkat-greedy-bisim}, we introduce \emph{symbolic GKAT coalgebra}, and lift the greedy bisimulation algorithm to the symbolic setting.
    \item In~\Cref{sec:symb-gkat-construction}, we lift the Thompson's construction and derivative construction to the symbolic setting.
    Then we establish their connection via a coalgebra homomorphism which allows us to derive the correctness and finiteness of one construction from the other.
    \item Finally, in~\Cref{sec:implementation}, we describe our implementation of symbolic derivative, symbolic Thompson's construction, and symbolic greedy bisimulation.
    We analyze algorithm performance under different satisfiability solver backends, and compare our work to previous implementations.
    When using MiniSat~\cite{een_MINISAT_2004} as the backend, our implementation can solve most test cases in less than a second with minimal memory usage, whereas previous implementation~\cite{pous_SymbolicAlgorithmsLanguage_2015} almost always result in a timeout.
\end{itemize}

\section{Background on Coalgebra and GKAT}\label{sec:background}
